---
title: "Ecosystem Forecasting Project Report"
format:
  html:
    embed-resources: true
---

Maria Popescu

#Part 1 – Model Description

##Model structure and rationale
-My Forecast is predicting Deep Chlorophyll Maxima depth at FCR for the VERA forecasting challenge. I am choosing to forecast at FCR only because BVR recently had a draw-down that leads to some complexity when navigating training the model. BVR water levels have changed substantially while FCR water levels have not.
-For my covariates, I am using Schmidt stability (derived from FLARE temperature profiles and FCR bathymetry), secchi depth (from FLARE), air temperature (from NOAA), wind speed (from NOAA using eastward and northward wind magnitude), and doy_sin/doy_cos to include seasonality. I only have driver uncertainty from my weather covariates (150 parameters). 
-I am using random forest for the model, and the hyperparameters are selected using a grid search with the best OOB r2 parameters being used for the final model. I split my training and testing data 80/20, and the residual standard deviation from the test data is used to include process uncertainty.

##Explanation of substantive novelty
-My modelling approach is different than what we have developed in class together thus far. I am using random forest, and grid search hyperparameter tuning. I am also using FLARE output to train and predict my observations and am using new covariates (one of which I have to calculate within the model. I calculate schmidt stability from the bathymetry data and temperature profile output from FLARE).

##Uncertainty sources and implementation
-My forecasts produce 150 ensemeble members to provide uncertainty. Though the FLARE output (schmidt stability and secchi depth) is summarized to a daily mean to reduce driver uncertainty, the weather ensembles are still cycled through to produce 150 ensemble members and integrated in the forecast to provide driver uncertainty. I also have process uncertainty from the test data sd of residuals which is added iteratively to each forecast. 

##Responses from the registration form
NEED TO COMPLETE 


#Part 2 – Reforecast Analysis
Definition
A reforecast mimics real‑time forecasting for a past period.
Required Properties
Must span at least 1 year
Each forecast must use only data available at its reference_datetime
No future observations
No observed covariates beyond the reference_datetime
Only driver forecasts that would have been available at the time
Data leakage invalidates the reforecast.
Do not use the submit function in your reforecast code to avoid submission to the Challenge.

Baseline Comparisons
Baseline scores (climatology and persistence) are accessed directly from the challenge catalog; you do not generate them yourself.

Performance evaluated across multiple axes:
Forecast horizon
Sites
Season or time of year
Comparison with baseline models
Written interpretation of results

1.  Accessing climatology, persistence, and using MY MODEL for DCM depth
```{R}
library(tidyverse)
library(arrow)
library(kableExtra)
library(scoringRules)
library(duckdbfs)

#updated
targets <- read_csv("https://amnh1.osn.mghpcc.org/bio230121-bucket01/vera4cast/targets/project_id=vera4cast/duration=P1D/daily-insitu-targets.csv.gz")
```
#Downloading all the forecasts generated by climatology and persistence.
I am going to do from January 2025-January 2026

```{r}
all_forecasts_full <- duckdbfs::open_dataset(
  "s3://anonymous@bio230121-bucket01/vera4cast/forecasts/bundled-summaries/project_id=vera4cast?endpoint_override=amnh1.osn.mghpcc.org",
  anonymous = TRUE
) |>
  filter(reference_datetime >= as_datetime("2025-01-01 00:00:00"),
         reference_datetime <= as_datetime("2026-01-01 00:00:00"),
         site_id == "fcre",
         model_id %in% c("climatology", "persistenceRW")) |> 
  collect() 

#to be able to plot performance over the course of horizon I need a column for horizon


#write_csv(all_forecasts_full, "all_forecasts_full.csv")
```
#Calculating CRPS for climatology, persistence, and my model for all reference datetimes and horizons. Which model has the lower score when you consider all submissions?

```{r}
target_variable <- "ChlorophyllMaximum_depth_sample"
focal_site <- "fcre"

obs_all <- targets |>
  mutate(datetime = as_date(datetime)) |>
  filter(variable == target_variable, site_id == focal_site) |>
  select(datetime, site_id, observation)

# Climatology (normal)
#only goes from 2025-01-08 - 2026-01-01
#MY DATES DO NOT ALIGN AT ALL WITH THE OBSERVATIONS SO THIS COMES BACK EMPTY BECAUSE OF THE INNERJOIN
clim_all <- all_forecasts_full |>
  filter(model_id == "climatology",
         variable == target_variable) |>
  inner_join(obs_all, by = c("datetime", "site_id")) |>
  filter(!is.na(observation), !is.na(mean), !is.na(sd)) |>
  mutate(crps = crps_norm(y = observation, mean = mean, sd = sd)) |>
  select(reference_datetime, datetime, site_id, variable, model_id, crps)
#family is normal here

# Persistence models
#family is also normal here
pers_all <- all_forecasts_full |>
  filter(model_id == "persistenceRW",
         variable == target_variable) |>
  inner_join(obs_all, by = c("datetime", "site_id")) |>
  filter(!is.na(observation), !is.na(mean), !is.na(sd)) |>
  mutate(crps = crps_norm(y = observation, mean = mean, sd = sd)) |>
  select(reference_datetime, datetime, site_id, variable, model_id, crps)
```

#Adding in my forecast

```{r}
#| include: false

all_forecast_dates <- seq(from = ymd("2025-04-01"),
                          to   = ymd("2026-04-01"),
                          by   = "day")

my_forecasts_full <- NULL
# Load helper functions
helper.functions <- source("./R/helper_functions.R")
#from mp_DCMdepth_Forecast.R
sites <- c("fcre") #just FCR
forecast_depths <- 'focal'

forecast_horizon <- 31
n_members <- 150
calibration_start_date <- ymd("2022-11-11")
model_id <- "fDCMdepth_mp"
targets_url <- "https://amnh1.osn.mghpcc.org/bio230121-bucket01/vera4cast/targets/project_id=vera4cast/duration=P1D/daily-insitu-targets.csv.gz"

var <- "ChlorophyllMaximum_depth_sample"
project_id <- "vera4cast"


for (fdate in all_forecast_dates) {
  
  forecast_date <- as_date(fdate)

  reforecast_output <- generate_DCMdepth_forecast(forecast_date = forecast_date, #changed this here to loop through
                                                forecast_horizon = forecast_horizon,
                                                n_members = n_members,
                                                output_folder = output_folder,
                                                model_id = model_id,
                                                targets_url = targets_url,
                                                var = var,
                                                site = sites,
                                                forecast_depths = NA_real_,
                                                project_id = project_id,
                                                calibration_start_date = calibration_start_date,
                                                seed = 'yes')

  my_forecasts_full <- bind_rows(my_forecasts_full, reforecast_output)}
  

```


```{r}
#| include: false
#from reforecasting assignment
mine_all <- my_forecasts_full |>
  inner_join(obs_all, by = "datetime") |>
  filter(!is.na(observation), !is.na(prediction)) |>
  group_by(reference_datetime, datetime, model_id) |>
  summarise(crps = crps_sample(y = unique(observation), dat = prediction),
            .groups = "drop")

crps_full <- bind_rows(clim_all, pers_all, mine_all)|>
  mutate(horizon = as.numeric(difftime(datetime, reference_datetime, units = "days"))) #for the horizon

mean_crps_by_horizon <- crps_full |>
  group_by(model_id, horizon) |>
  summarise(mean_crps = mean(crps, na.rm = TRUE),
            n = n(),
            .groups = "drop") |>
  arrange(mean_crps)

```

#mean crps by horizon
```{r}
ggplot(mean_crps_by_horizon, aes(x = horizon, y = mean_crps, color = model_id))+
  geom_line() +
  geom_point() +
  labs(x = "horizon", y = "CRPS", color = "model")

```
#could also do by date

```{r}
mean_crps_by_date <- crps_full |>
  group_by(model_id, reference_datetime) |> #probs this isn't right putting this here for now till model finishes running so I can check the column
  summarise(mean_crps = mean(crps, na.rm = TRUE),
            n = n(),
            .groups = "drop") |>
  arrange(mean_crps)

ggplot(mean_crps_by_horizon, aes(x = horizon, y = mean_crps, color = model_id))+
  geom_line() +
  geom_point() +
  labs(x = "horizon", y = "CRPS", color = "model")

```


#written interpretation of results




#Part 3 – Reproducibility Checklist
Include:
model_id: "fDCMdepth_mp"
GitHub repository link (this is a new repo): https://github.com/MariaPopescu123/Maria_Forecasting_Project  

Link to forecast‑generation code: https://github.com/MariaPopescu123/Maria_Forecasting_Project/blob/main/Submit_forecast/mp_DCMdepth_Forecast.R

Link to Quarto (.qmd) report: https://github.com/MariaPopescu123/Maria_Forecasting_Project/blob/main/Project_Report.qmd

Evidence of multiple automated forecasts:

Link to model entry in the challenge catalog:
